---
title: "South Africa Covid19"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

INTRODUCTION
www.corona-stats.co.za launched a free API for historical South African COVID-19 data in April 2020. Data can be accessed using the link:"http://corona-stats.mobi/api/. This project entails accessing South African COVID-19 data from the API, compiling accessed data into useable datasets,Exploratory data analysis,data cleaning and representing the data in useable format i.e. data visualization utilizing Tableau.

1. DATA ACCESS
To extract data from the API R programming language packages:httr and Jsonlite were used.

```{r}
library(httr)
library(jsonlite)
library(dplyr)

base_url = "http://corona-stats.mobi/api/"
info_url = "json.2.0.php?key=b1fQnVw3MoTAyqNmkigJ"
full_url = base::paste0(base_url,info_url)

api_call = httr::GET(full_url)

```

To check whether the request was successful the status code has to be 200. If it is around 400 it means the request was unsuccessful.
```{r}
api_call$status_code
```
Converting the response to usable format
```{r}
api_char = base::rawToChar(api_call$content)

str(api_char)

api_Json = jsonlite::fromJSON(api_char,flatten = T)

api_Json
str(api_Json)
Covid19_RSA = api_Json$RSA
```
The Covid19 data is in the RSA list.RSA contains 22 lists of South Africa Covid19 data. 

2.COMPILING ACCESSED DATA TO USEABLE DATASETS
a.National Data
```{r}
Dates1 = Covid19_RSA$Dates
View(Dates1)
SA_data = Covid19_RSA$National
Cases = SA_data$Cases
Recoveries = SA_data$Recoveries
Deaths = SA_data$Deaths
Tests = SA_data$Tests
Vaccinations = SA_data$Vaccinations
Daily_Cases = SA_data$`Daily Cases`
Daily_Deaths = SA_data$`Daily Deaths`
Daily_Vaccinations = SA_data$`Daily Vaccinations`
Cases_14day_average = SA_data$`Daily Cases 14 day average`
Deaths_14day_average = SA_data$`Daily Deaths 14 day average`
Vaccinations_14day_average = SA_data$`Vaccinations 14 day average`

mySA_Covid19data = data.frame(Date=Dates1,Total_Cases=Cases,Total_Recoveries=Recoveries,
                              Total_Deaths=Deaths,Total_Tests=Tests,Total_Vaccinations=Vaccinations,
                              Daily_Cases,Daily_Deaths,Daily_Vaccinations,Cases_14day_average,
                              Deaths_14day_average,Vaccinations_14day_average)

View(mySA_Covid19data)

library(writexl)

write_xlsx(mySA_Covid19data,"SaCovid19.xlsx")

```

b. Western Cape Data
```{r}
WC_data = Covid19_RSA$WC
Total_Cases = WC_data$Cases
Total_Recoveries = WC_data$Recoveries
Total_Deaths = WC_data$Deaths

Daily_Cases = WC_data$`Daily Cases`
Daily_Deaths = WC_data$`Daily Deaths`

Cases_14day_average = WC_data$`Daily Cases 14 day average`
Deaths_14day_average = WC_data$`Daily Deaths 14 day average`


WC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(WC_Covid19data,"WesternCapeCovid19.xlsx")

```

c. Gauteng Data
```{r}
Gauteng_data = Covid19_RSA$GP
Total_Cases = Gauteng_data$Cases
Total_Recoveries = Gauteng_data$Recoveries
Total_Deaths = Gauteng_data$Deaths

Daily_Cases = Gauteng_data$`Daily Cases`
Daily_Deaths = Gauteng_data$`Daily Deaths`

Cases_14day_average = Gauteng_data$`Daily Cases 14 day average`
Deaths_14day_average = Gauteng_data$`Daily Deaths 14 day average`


Gauteng_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(Gauteng_Covid19data,"GautengCovid19.xlsx")
```

d. KZN Data
```{r}
KZN_data = Covid19_RSA$KZN
Total_Cases = KZN_data$Cases
Total_Recoveries = KZN_data$Recoveries
Total_Deaths = KZN_data$Deaths

Daily_Cases = KZN_data$`Daily Cases`
Daily_Deaths = KZN_data$`Daily Deaths`

Cases_14day_average = KZN_data$`Daily Cases 14 day average`
Deaths_14day_average = KZN_data$`Daily Deaths 14 day average`


KZN_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(KZN_Covid19data,"KZNCovid19.xlsx")

```

e. Eastern Cape Data
```{r}
EC_data = Covid19_RSA$EC
Total_Cases = EC_data$Cases
Total_Recoveries = EC_data$Recoveries
Total_Deaths = EC_data$Deaths

Daily_Cases = EC_data$`Daily Cases`
Daily_Deaths = EC_data$`Daily Deaths`

Cases_14day_average = EC_data$`Daily Cases 14 day average`
Deaths_14day_average = EC_data$`Daily Deaths 14 day average`


EC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(EC_Covid19data,"ECCovid19.xlsx")
```

f. Free State Data
```{r}
FS_data = Covid19_RSA$FS
Total_Cases = FS_data$Cases
Total_Recoveries = FS_data$Recoveries
Total_Deaths = FS_data$Deaths

Daily_Cases = FS_data$`Daily Cases`
Daily_Deaths = FS_data$`Daily Deaths`

Cases_14day_average = FS_data$`Daily Cases 14 day average`
Deaths_14day_average = FS_data$`Daily Deaths 14 day average`


FS_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(FS_Covid19data,"FSCovid19.xlsx")
```

g. Mpumalanga Data
```{r}
MP_data = Covid19_RSA$MP
Total_Cases = MP_data$Cases
Total_Recoveries = MP_data$Recoveries
Total_Deaths = MP_data$Deaths

Daily_Cases = MP_data$`Daily Cases`
Daily_Deaths = MP_data$`Daily Deaths`

Cases_14day_average = MP_data$`Daily Cases 14 day average`
Deaths_14day_average = MP_data$`Daily Deaths 14 day average`


MP_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(MP_Covid19data,"MPCovid19.xlsx")
```

h. Limpopo Data
```{r}
LP_data = Covid19_RSA$LP
Total_Cases = LP_data$Cases
Total_Recoveries = LP_data$Recoveries
Total_Deaths = LP_data$Deaths

Daily_Cases = LP_data$`Daily Cases`
Daily_Deaths = LP_data$`Daily Deaths`

Cases_14day_average = LP_data$`Daily Cases 14 day average`
Deaths_14day_average = LP_data$`Daily Deaths 14 day average`


LP_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(LP_Covid19data,"LPCovid19.xlsx")
```

i. North West Data
```{r}
NW_data = Covid19_RSA$NW
Total_Cases = NW_data$Cases
Total_Recoveries = NW_data$Recoveries
Total_Deaths = NW_data$Deaths

Daily_Cases = NW_data$`Daily Cases`
Daily_Deaths = NW_data$`Daily Deaths`

Cases_14day_average = NW_data$`Daily Cases 14 day average`
Deaths_14day_average = NW_data$`Daily Deaths 14 day average`


NW_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(NW_Covid19data,"NWCovid19.xlsx")

```

j. Northern Cape Data
```{r}
NC_data = Covid19_RSA$NC
Total_Cases = NC_data$Cases
Total_Recoveries = NC_data$Recoveries
Total_Deaths = NC_data$Deaths

Daily_Cases = NC_data$`Daily Cases`
Daily_Deaths = NC_data$`Daily Deaths`

Cases_14day_average = NC_data$`Daily Cases 14 day average`
Deaths_14day_average = NC_data$`Daily Deaths 14 day average`


NC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

write_xlsx(NC_Covid19data,"NCCovid19.xlsx")

```

All the provincial dataframes were combined into 1 dataset, Sa_ProvincialCOVID19.


2. EXPLORATORY DATA ANALYSIS
```{r}
mydata = Sa_ProvincialCOVID19
```

A. Checking the total number of rows and columns
```{r}
nrow(mydata)
ncol(mydata)
```
B. Basic data profiling
```{r}
library(funModeling)
df_status(mydata)
```
Columns q_zeros and p_zeros show the number of zeros and percentage of zeros per column respectively. Columns q_na and p_na show the number of na values and percentage of na values respectively.Column type shows the data types for all the columns. Some numerical values are read as characters above and also the date format needs to be changed.
```{r}
mydata$Date = as.Date(mydata$Date)
mydata$Daily_Deaths = as.numeric(mydata$Daily_Deaths)
mydata$Total_Cases = as.numeric(mydata$Total_Cases)
mydata$Total_Recoveries = as.numeric(mydata$Total_Recoveries)
mydata$Total_Deaths = as.numeric(mydata$Total_Deaths)
mydata$Daily_Cases = as.numeric(mydata$Daily_Cases)
mydata$Cases_14day_average = as.numeric(mydata$Cases_14day_average)
mydata$Deaths_14day_average = as.numeric(mydata$Deaths_14day_average)

df_status(mydata)
```
All the column data types have been corrected.
To find the statistical properties of the data:
```{r}
summary(mydata)
```
There appears to be an error in the Daily_Cases and Deaths_14day_average columns. The minimum values are negative and that needs to be rectified. All the negative numbers will be replaced by the variable's absolute value.
```{r}
mydata_sorted = mydata[order(mydata$Date),]
mydata_sorted$Daily_Cases = abs(mydata_sorted$Daily_Cases)
mydata_sorted$Deaths_14day_average = abs(mydata_sorted$Deaths_14day_average)
summary(mydata_sorted)

```
A. UNIVARIATE ANALYSIS

i. Daily Cases
```{r}
summary(mydata_sorted$Daily_Cases)
par(mfrow=c(1,3))
hist(mydata_sorted$Daily_Cases, main = "Daily Cases",
     xlab = "Daily cases",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Daily_Cases), main = "Total Cases distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Daily_Cases,col = "blue",main="Boxplot:Daily Cases")


```
```{r}
summary(mydata_sorted$Daily_Cases)
```
The daily cases are skewed to the left as shown in the descriptive stats, histogram, density plot and boxplot. There appears to be outliers as shown by the boxplot but these values are extreme values which represent a time when COVID 19 cases were on the rise. The most daily cases observed in the data are between 0 and 1000.

```{r}
plot(mydata_sorted$Date,mydata_sorted$Daily_Cases,type = "l",lwd = 1, col = "#cc0000",main = "Daily COVID19 cases with time",
     xlab = "Date",
     ylab = "Daily COVID19 cases")

```
The first peak corresponds to the first COVID19 outbreak in May/June 2020. The second peak between end of 2020 and beginning of 2021 corresponds to the second outbreak. The peak in mid 2022 corresponds to the outbreak of the Delta variant, which has a higher transmissibility. Torwards the end of 2021 and beginning of 2022 there was an outbreak of the Omicron variant. This corresponds to the last peak in the plot above.


ii Daily Deaths
```{r}
par(mfrow=c(1,3))
mydata_sorted$Daily_Deaths = abs(mydata_sorted$Daily_Deaths)
mydata_sorted$Daily_Deaths = as.numeric(mydata_sorted$Daily_Deaths)
hist(mydata_sorted$Daily_Deaths, main = "Daily Deaths",
     xlab = "Daily deaths",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Daily_Deaths), main = "Total Deaths distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Daily_Deaths,col = "blue",main="Boxplot:Daily Deaths")

```

The daily deaths are skewed to the left as shown in the descriptive stats, histogram, density plot and boxplot. There appears to be outliers as shown by the boxplot but these values are extreme values which represent a time when COVID 19 deaths were on the rise and this corresponds to the characteristics of the COVID19 Daily Cases. Most days recorded daily deaths between 0 and 50. The maximum number of deaths recorded a day is 515.
```{r}
summary(mydata_sorted$Daily_Deaths)
```

```{r}
plot(mydata_sorted$Date,mydata_sorted$Daily_Deaths,type = "l",lwd = 1, col = "blue",main = "Daily COVID19 deaths with time",
     xlab = "Date",
     ylab = "Daily COVID19 deaths")
```

iii Total Recoveries
```{r}
par(mfrow=c(1,3))
hist(mydata_sorted$Total_Recoveries, main = "Total Recoveries",
     xlab = "Total Recoveries",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Recoveries),main = "Total Recoveries", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Recoveries,col = "blue",main="Boxplot:Total Recoveries")

```

iv Total Cases
```{r}
par(mfrow=c(1,3))
hist(mydata_sorted$Total_Cases, main = "Total Cases",
     xlab = "Total Cases",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Cases), main = "Total Cases distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Cases,col = "blue",main="Boxplot:Total Cases")

```
v Total Deaths
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Total_Deaths, main = "Total Deaths",
     xlab = "Total deaths",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Deaths), main = "Total Deaths distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Deaths,col = "blue",main="Boxplot:Total Deaths")
```
vi Daily Cases 14 day average
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Cases_14day_average, main = "Cases:14day average",
     xlab = "cases:14day average",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Cases_14day_average), main = "Cases:14day average", col = "red",lwd = 2)
boxplot(mydata_sorted$Cases_14day_average,col = "blue",main="Boxplot:Cases 14day average")
```
vii Deaths 14 day average
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Deaths_14day_average, main = "Deaths:14day average",
     xlab = "Deaths 14day average",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Deaths_14day_average), main = "Deaths 14day average distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Deaths_14day_average,col = "blue",main="Boxplot:Deaths 14day average")

```


MULTIVARIATE ANALYSIS
```{r}
by(mydata_sorted$Total_Cases,mydata_sorted$Province,summary)
```

```{r}
by(mydata_sorted$Total_Deaths,mydata_sorted$Province,summary)
```
Though Gauteng has had the highest number of total COVID19 cases, more people have died due to COVID19 related complications in Western Cape.

```{r}
boxplot(mydata_sorted$Total_Cases~mydata_sorted$Province,notch = T,col="blue",main =
          "Boxplot:Total Cases per province")
```
```{r}
boxplot(mydata_sorted$Total_Deaths~mydata_sorted$Province,notch = T,col="#cc0000", main = 
          "Boxplots:Total Deaths per province")
```
The boxplots above give a relative comparison of the total cases and total deaths for all the provinces.

```{r}
plot(mydata_sorted$Total_Cases,mydata_sorted$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
There is a direct correlation between the total cases and total deaths.Different lines shown in the scatter plot represents 10 different provinces.

Scatter plots: Total cases vs Total deaths per province
1. Gauteng
```{r}
gauteng = subset(mydata_sorted,Province == "Gauteng")
View(gauteng)
```

```{r}
plot(gauteng$Total_Cases,gauteng$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Gauteng:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
We can define death rate as the number of deaths divided by the number of cases.In this case we took the maximum total of deaths divided by the maximum total cases. The death rate is 0.0170. This variable enables a comparison of the number of deaths as per number of cases to be compared province by province. This can probably give an indication of the quality of care for covid19 patients across provinces. For Gauteng the death rate is 0.017. Even though Gauteng recorded the highest total infections and is currently second highest in total deaths but the number of deaths per cases is lower than other provinces for instance Eastern Cape which is: 0.048.

2. Eastern Cape
```{r}
EC = subset(mydata_sorted,Province == "Eastern Cape")
plot(EC$Total_Cases,EC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Eastern Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

3. Free State
```{r}
FS = subset(mydata_sorted,Province == "Free State")
plot(FS$Total_Cases,FS$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Free State:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")


```
Death rate = 0.0378

4. KwaZulu-Natal
```{r}
kzn = subset(mydata_sorted,Province == "KwaZulu-Natal")
plot(kzn$Total_Cases,kzn$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "KwaZulu-Natal:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
Death rate = 0.0243

5. Mpumalanga
```{r}
MP = subset(mydata_sorted,Province == "Mpumalanga")
plot(MP$Total_Cases,MP$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Mpumalanga:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
Death rate = 0.0243

6. Limpopo
```{r}
LP = subset(mydata_sorted,Province == "Limpopo")
plot(LP$Total_Cases,LP$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Limpopo:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
Death rate = 0.0294

7. North West
```{r}
NW = subset(mydata_sorted,Province == "North West")
plot(NW$Total_Cases,NW$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "North West:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
Death rate = 0.024

8. Northern Cape
```{r}
NC = subset(mydata_sorted,Province == "Northern Cape")
plot(NC$Total_Cases,NC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Northern Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```
Death rate = 0.0295

9. Western Cape
```{r}
WC = subset(mydata_sorted,Province == "Western Cape")
plot(WC$Total_Cases,WC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Western Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")

```
Death rate = 0.0336

3. DATA VISUALIZATION USING TABLEAU

(i)Total confirmed cases and deaths over time dashboard:
https://public.tableau.com/app/profile/mufaro.chinake/viz/SouthAfricaCOVID19ConfirmedCases/Dashboard1

(ii) Daily cases and deaths over time
https://public.tableau.com/app/profile/mufaro.chinake/viz/SouthAfricaCOVID19DailyCasesandDeaths/Dashboard1


