---
title: "South Africa Covid19 Data Analysis"
output:
   html_document:
        number_sections: true
        toc: true
        toc_float: true
        toc_depth: 2
        theme: journal
---

<h1>1.Introduction</h1>
<p>
www.corona-stats.co.za launched a free API for historical South African
COVID-19 data in April 2020. Data can be accessed using the
link:"<http://corona-stats.mobi/api/>. This project entails accessing
South African COVID-19 data from the API, compiling accessed data into
useable datasets,Exploratory data analysis,data cleaning and
representing the data in useable format i.e. data visualization
utilizing Tableau.
</p>
<p>
The chunks of code below were used to access and compile data from the API. When
report was knitted from RMarkdown the API key had expired. The data used inthe 
analysis was accessed and compiled before they key expired.
</p>

<h1>2.Data access </h1>
To extract data from the API R programming language packages:httr and Jsonlite were used.
```{r setup,cache=FALSE}
knitr::opts_chunk$set(error = TRUE)
```


```{r}
library(httr)
library(jsonlite)
library(dplyr)

base_url = "http://corona-stats.mobi/api/"
info_url = "json.2.0.php?key=J7t2b0NhyD8PSqT9ZMEY"
full_url = base::paste0(base_url,info_url)

api_call = httr::GET(full_url)


```

To check whether the request was successful the status code has to be
200. If it is around 400 it means the request was unsuccessful.

```{r}
api_call$status_code
```

Converting the response to usable format

```{r}
api_char = base::rawToChar(api_call$content)

str(api_char)

api_Json = jsonlite::fromJSON(api_char,flatten = T)

api_Json
str(api_Json)
Covid19_RSA = api_Json$RSA
```

The Covid19 data is in the RSA list.RSA contains 22 lists of South
Africa Covid19 data.

<h1>3.Compiling accessed data to useable datasets</h1>

<h2>a.National Data</h2>
```{r}
Dates1 = Covid19_RSA$Dates
View(Dates1)
SA_data = Covid19_RSA$National
Cases = SA_data$Cases
Recoveries = SA_data$Recoveries
Deaths = SA_data$Deaths
Tests = SA_data$Tests
Vaccinations = SA_data$Vaccinations
Daily_Cases = SA_data$`Daily Cases`
Daily_Deaths = SA_data$`Daily Deaths`
Daily_Vaccinations = SA_data$`Daily Vaccinations`
Cases_14day_average = SA_data$`Daily Cases 14 day average`
Deaths_14day_average = SA_data$`Daily Deaths 14 day average`
Vaccinations_14day_average = SA_data$`Vaccinations 14 day average`

mySA_Covid19data = data.frame(Date=Dates1,Total_Cases=Cases,Total_Recoveries=Recoveries,
                              Total_Deaths=Deaths,Total_Tests=Tests,Total_Vaccinations=Vaccinations,
                              Daily_Cases,Daily_Deaths,Daily_Vaccinations)

View(mySA_Covid19data)

library(writexl)

write_xlsx(mySA_Covid19data,"SaCovid19.xlsx")

```

<h2>b. Western Cape Data</h2>
```{r}
WC_data = Covid19_RSA$WC
Total_Cases = WC_data$Cases
Total_Recoveries = WC_data$Recoveries
Total_Deaths = WC_data$Deaths

Daily_Cases = WC_data$`Daily Cases`
Daily_Deaths = WC_data$`Daily Deaths`

Cases_14day_average = WC_data$`Daily Cases 14 day average`
Deaths_14day_average = WC_data$`Daily Deaths 14 day average`


WC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(WC_Covid19data,"WesternCapeCovid19.xlsx")
library(readxl)
#WC = read_excel("~/GitHub/South-Africa-COVID19/WesternCapeCovid19.xlsx")
#WC_data = mutate(WC,Province = "Western Cape")

```

<h2>c. Gauteng Data</h2>
```{r}
Gauteng_data = Covid19_RSA$GP
Total_Cases = Gauteng_data$Cases
Total_Recoveries = Gauteng_data$Recoveries
Total_Deaths = Gauteng_data$Deaths

Daily_Cases = Gauteng_data$`Daily Cases`
Daily_Deaths = Gauteng_data$`Daily Deaths`

Cases_14day_average = Gauteng_data$`Daily Cases 14 day average`
Deaths_14day_average = Gauteng_data$`Daily Deaths 14 day average`


Gauteng_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(Gauteng_Covid19data,"GautengCovid19.xlsx")

#Gauteng = read_excel("~/GitHub/South-Africa-COVID19/GautengCovid19.xlsx")
#Gauteng_data = mutate(Gauteng,Province = "Gauteng")
```

<h2>d. KZN Data</h2>
```{r}
KZN_data = Covid19_RSA$KZN
Total_Cases = KZN_data$Cases
Total_Recoveries = KZN_data$Recoveries
Total_Deaths = KZN_data$Deaths

Daily_Cases = KZN_data$`Daily Cases`
Daily_Deaths = KZN_data$`Daily Deaths`

Cases_14day_average = KZN_data$`Daily Cases 14 day average`
Deaths_14day_average = KZN_data$`Daily Deaths 14 day average`


KZN_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(KZN_Covid19data,"KZNCovid19.xlsx")

#KZN = read_excel("~/GitHub/South-Africa-COVID19/KZNCovid19.xlsx")
#KZN_data = mutate(KZN,Province = "KwaZulu-Natal")

```

<h2>e. Eastern Cape Data</h2>
```{r}
EC_data = Covid19_RSA$EC
Total_Cases = EC_data$Cases
Total_Recoveries = EC_data$Recoveries
Total_Deaths = EC_data$Deaths

Daily_Cases = EC_data$`Daily Cases`
Daily_Deaths = EC_data$`Daily Deaths`

Cases_14day_average = EC_data$`Daily Cases 14 day average`
Deaths_14day_average = EC_data$`Daily Deaths 14 day average`


EC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(EC_Covid19data,"ECCovid19.xlsx")

#EC = read_excel("~/GitHub/South-Africa-COVID19/ECCovid19.xlsx")
#EC_data = mutate(EC,Province = "Eastern Cape")
```

<h2>f. Free State Data</h2>
```{r}
FS_data = Covid19_RSA$FS
Total_Cases = FS_data$Cases
Total_Recoveries = FS_data$Recoveries
Total_Deaths = FS_data$Deaths

Daily_Cases = FS_data$`Daily Cases`
Daily_Deaths = FS_data$`Daily Deaths`

Cases_14day_average = FS_data$`Daily Cases 14 day average`
Deaths_14day_average = FS_data$`Daily Deaths 14 day average`


FS_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(FS_Covid19data,"FSCovid19.xlsx")

#FS = read_excel("~/GitHub/South-Africa-COVID19/FSCovid19.xlsx")
#FS_data = mutate(FS,Province = "Free State")
```

<h2>g. Mpumalanga Data</h2>
```{r}
MP_data = Covid19_RSA$MP
Total_Cases = MP_data$Cases
Total_Recoveries = MP_data$Recoveries
Total_Deaths = MP_data$Deaths

Daily_Cases = MP_data$`Daily Cases`
Daily_Deaths = MP_data$`Daily Deaths`

Cases_14day_average = MP_data$`Daily Cases 14 day average`
Deaths_14day_average = MP_data$`Daily Deaths 14 day average`


MP_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(MP_Covid19data,"MPCovid19.xlsx")

#MP = read_excel("~/GitHub/South-Africa-COVID19/MPCovid19.xlsx")
#MP_data = mutate(MP,Province = "Mpumalanga")
```

<h2>h. Limpopo Data</h2>
```{r}
LP_data = Covid19_RSA$LP
Total_Cases = LP_data$Cases
Total_Recoveries = LP_data$Recoveries
Total_Deaths = LP_data$Deaths

Daily_Cases = LP_data$`Daily Cases`
Daily_Deaths = LP_data$`Daily Deaths`

Cases_14day_average = LP_data$`Daily Cases 14 day average`
Deaths_14day_average = LP_data$`Daily Deaths 14 day average`


LP_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(LP_Covid19data,"LPCovid19.xlsx")

#LP = read_excel("~/GitHub/South-Africa-COVID19/LPCovid19.xlsx")
#LP_data = mutate(LP, Province = "Limpopo")
```

<h2>i. North West Data</h2>
```{r}
NW_data = Covid19_RSA$NW
Total_Cases = NW_data$Cases
Total_Recoveries = NW_data$Recoveries
Total_Deaths = NW_data$Deaths

Daily_Cases = NW_data$`Daily Cases`
Daily_Deaths = NW_data$`Daily Deaths`

Cases_14day_average = NW_data$`Daily Cases 14 day average`
Deaths_14day_average = NW_data$`Daily Deaths 14 day average`


NW_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(NW_Covid19data,"NWCovid19.xlsx")
#NW = read_excel("~/GitHub/South-Africa-COVID19/NWCovid19.xlsx")
#NW_data = mutate(NW,Province = "North West")

```

<h2>j. Northern Cape Data</h2>
```{r}
NC_data = Covid19_RSA$NC
Total_Cases = NC_data$Cases
Total_Recoveries = NC_data$Recoveries
Total_Deaths = NC_data$Deaths

Daily_Cases = NC_data$`Daily Cases`
Daily_Deaths = NC_data$`Daily Deaths`

Cases_14day_average = NC_data$`Daily Cases 14 day average`
Deaths_14day_average = NC_data$`Daily Deaths 14 day average`


NC_Covid19data = data.frame(Date=Dates1,Total_Cases,Total_Recoveries,Total_Deaths,Daily_Cases,Daily_Deaths,Cases_14day_average,Deaths_14day_average)

#write_xlsx(NC_Covid19data,"NCCovid19.xlsx")

#NC = read_excel("~/GitHub/South-Africa-COVID19/NCCovid19.xlsx")
#NC_data = mutate(NC,Province = "Northern Cape")
```

All the provincial dataframes were combined into 1 dataset,
Sa_ProvincialCOVID19.

<h1>4. Exploratory Data Analysis</h1>

```{r}
Sa_ProvincialCOVID19 = rbind(WC_data,KZN_data,Gauteng_data,EC_data,FS_data,MP_data,LP_data,NW_data,NC_data)

Sa_ProvincialCOVID19 = read_excel("~\\GitHub\\South-Africa-COVID19\\docs\\Sa_ProvincialCovid19.xlsx")
mydata = Sa_ProvincialCOVID19
```

<h2>A. Checking the total number of rows and columns</h2>

```{r}
nrow(mydata)
ncol(mydata)
```

<h2>B. Basic data profiling</h2>

```{r}
library(funModeling)
df_status(mydata)
```

Columns q_zeros and p_zeros show the number of zeros and percentage of
zeros per column respectively. Columns q_na and p_na show the number of
na values and percentage of na values respectively.Column type shows the
data types for all the columns. Some numerical values are read as
characters above and also the date format needs to be changed.

```{r}
mydata$Date = as.Date(mydata$Date)
mydata$Daily_Deaths = as.numeric(mydata$Daily_Deaths)
mydata$Total_Cases = as.numeric(mydata$Total_Cases)
mydata$Total_Recoveries = as.numeric(mydata$Total_Recoveries)
mydata$Total_Deaths = as.numeric(mydata$Total_Deaths)
mydata$Daily_Cases = as.numeric(mydata$Daily_Cases)
mydata$Cases_14day_average = as.numeric(mydata$Cases_14day_average)
mydata$Deaths_14day_average = as.numeric(mydata$Deaths_14day_average)

df_status(mydata)
```

All the column data types have been corrected. To find the statistical
properties of the data:

```{r}
summary(mydata)
```

There appears to be an error in the Daily_Deaths,Daily_Cases,
Deaths_14day_average and Cases_14day_average columns. The minimum values are negative and that needs to be rectified. All the negative numbers will be replaced by the
variable's absolute value.

```{r}
mydata_sorted = mydata[order(mydata$Date),]
mydata_sorted$Daily_Cases = abs(mydata_sorted$Daily_Cases)
mydata_sorted$Daily_Deaths = abs(mydata_sorted$Daily_Deaths)
mydata_sorted$Deaths_14day_average = abs(mydata_sorted$Deaths_14day_average)
mydata_sorted$Cases_14day_average = abs(mydata_sorted$Cases_14day_average)
summary(mydata_sorted)

```

<h2>C. Univariate Analysis</h2>

<h3>i. Daily Cases</h3>

```{r}
summary(mydata_sorted$Daily_Cases)
par(mfrow=c(1,3))
hist(mydata_sorted$Daily_Cases, main = "Daily Cases",
     xlab = "Daily cases",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Daily_Cases), main = "Total Cases distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Daily_Cases,col = "blue",main="Boxplot:Daily Cases")


```

```{r}
summary(mydata_sorted$Daily_Cases)
```

The daily cases are skewed to the right as shown in the descriptive
stats, histogram, density plot and boxplot. There appears to be outliers
as shown by the boxplot but these values are extreme values which
represent a time when COVID 19 cases were on the rise. The most daily
cases observed in the data are between 0 and 1000.

```{r}
plot(mydata_sorted$Date,mydata_sorted$Daily_Cases,type = "l",lwd = 1, col = "#cc0000",main = "Daily COVID19 cases with time",
     xlab = "Date",
     ylab = "Daily COVID19 cases")

```
<p>
The first peak corresponds to the first COVID19 outbreak in May/June
2020. The second peak between end of 2020 and beginning of 2021
corresponds to the second outbreak. The peak in mid 2022 corresponds to
the outbreak of the Delta variant, which has a higher transmission rate.
Torwards the end of 2021 and beginning of 2022 there was an outbreak of
the Omicron variant. This corresponds to the fourth peak in the plot
above.
</p>
<h3>ii Daily Deaths</h3>
```{r}
par(mfrow=c(1,3))
mydata_sorted$Daily_Deaths = abs(mydata_sorted$Daily_Deaths)
mydata_sorted$Daily_Deaths = as.numeric(mydata_sorted$Daily_Deaths)
hist(mydata_sorted$Daily_Deaths, main = "Daily Deaths",
     xlab = "Daily deaths",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Daily_Deaths), main = "Total Deaths distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Daily_Deaths,col = "blue",main="Boxplot:Daily Deaths")

```

The daily deaths are skewed to the right as shown in the descriptive
stats, histogram, density plot and boxplot. There appears to be outliers
as shown by the boxplot but these values are extreme values which
represent a time when COVID 19 deaths were on the rise and this
corresponds to the characteristics of the COVID19 Daily Cases. Most days
recorded daily deaths between 0 and 50. The maximum number of deaths
recorded a day is 515.

```{r}
summary(mydata_sorted$Daily_Deaths)
```

```{r}
plot(mydata_sorted$Date,mydata_sorted$Daily_Deaths,type = "l",lwd = 1, col = "blue",main = "Daily COVID19 deaths with time",
     xlab = "Date",
     ylab = "Daily COVID19 deaths")
```

<h3>iii Total Recoveries</h3>
```{r}
par(mfrow=c(1,3))
hist(mydata_sorted$Total_Recoveries, main = "Total Recoveries",
     xlab = "Total Recoveries",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Recoveries),main = "Total Recoveries", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Recoveries,col = "blue",main="Boxplot:Total Recoveries")

```

<h3>iv Total Cases</h3>
```{r}
par(mfrow=c(1,3))
hist(mydata_sorted$Total_Cases, main = "Total Cases",
     xlab = "Total Cases",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Cases), main = "Total Cases distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Cases,col = "blue",main="Boxplot:Total Cases")

```

<h3>v.Total Deaths</h3>
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Total_Deaths, main = "Total Deaths",
     xlab = "Total deaths",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Total_Deaths), main = "Total Deaths distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Total_Deaths,col = "blue",main="Boxplot:Total Deaths")
```

<h3>vi.Cases 14 day average</h3>
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Cases_14day_average, main = "Cases:14day average",
     xlab = "cases:14day average",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Cases_14day_average), main = "Cases:14day average", col = "red",lwd = 2)
boxplot(mydata_sorted$Cases_14day_average,col = "blue",main="Boxplot:Cases 14day average")

```

<h3>vii.Deaths 14 day average</h3>
```{r}
par(mfrow=c(1,3))

hist(mydata_sorted$Deaths_14day_average, main = "Deaths:14day average",
     xlab = "Deaths 14day average",
     ylab = "Frequency",
     col = "blue")
plot(density(mydata_sorted$Deaths_14day_average), main = "Deaths 14day average distribution", col = "red",lwd = 2)
boxplot(mydata_sorted$Deaths_14day_average,col = "blue",main="Boxplot:Deaths 14day average")

```

<h2>D.Multivariate Analysis</h2>
```{r}
by(mydata_sorted$Total_Cases,mydata_sorted$Province,summary)
```

```{r}
by(mydata_sorted$Total_Deaths,mydata_sorted$Province,summary)
```

Though Gauteng has had the highest number of total COVID19 cases, more
people have died due to COVID19 related complications in Western Cape.

```{r}
boxplot(mydata_sorted$Total_Cases~mydata_sorted$Province,notch = T,col="blue",main =
          "Boxplot:Total Cases per province")
```

```{r}
boxplot(mydata_sorted$Total_Deaths~mydata_sorted$Province,notch = T,col="#cc0000", main = 
          "Boxplots:Total Deaths per province")
```

The boxplots above give a relative comparison of the total cases and
total deaths for all the provinces.

```{r}
plot(mydata_sorted$Total_Cases,mydata_sorted$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

There is a direct correlation between the total cases and total
deaths.Different lines shown in the scatter plot represents 10 different
provinces.

<h3>Scatter plots: Total cases vs Total deaths per province</h3>
<h4>1.Gauteng</h4>
```{r}
gauteng = subset(mydata_sorted,Province == "Gauteng")

```

```{r}
plot(gauteng$Total_Cases,gauteng$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Gauteng:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

We can define death rate as the number of deaths divided by the number
of cases.In this case we took the maximum total of deaths divided by the
maximum total cases. The death rate is 0.0170. This variable enables a
comparison of the number of deaths as per number of cases to be compared
province by province. This can probably give an indication of the
quality of care for covid19 patients across provinces. For Gauteng the
death rate is 0.017. Even though Gauteng recorded the highest total
infections and is currently second highest in total deaths but the
number of deaths per cases is lower than other provinces for instance
Eastern Cape which is: 0.048.

<h4>2. Eastern Cape</h4>
```{r}
EC = subset(mydata_sorted,Province == "Eastern Cape")
plot(EC$Total_Cases,EC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Eastern Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

<h4>3. Free State</h4>
```{r}
FS = subset(mydata_sorted,Province == "Free State")
plot(FS$Total_Cases,FS$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Free State:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")


```

Death rate = 0.0378

<h4>4. KwaZulu-Natal</h4>
```{r}
kzn = subset(mydata_sorted,Province == "KwaZulu-Natal")
plot(kzn$Total_Cases,kzn$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "KwaZulu-Natal:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

Death rate = 0.0243

<h4>5. Mpumalanga</h4>
```{r}
MP = subset(mydata_sorted,Province == "Mpumalanga")
plot(MP$Total_Cases,MP$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Mpumalanga:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

Death rate = 0.0243

<h4>6. Limpopo</h4>
```{r}
LP = subset(mydata_sorted,Province == "Limpopo")
plot(LP$Total_Cases,LP$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Limpopo:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

Death rate = 0.0294

<h4>7. North West</h4>
```{r}
NW = subset(mydata_sorted,Province == "North West")
plot(NW$Total_Cases,NW$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "North West:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

Death rate = 0.024

<h4>8. Northern Cape</h4>
```{r}
NC = subset(mydata_sorted,Province == "Northern Cape")
plot(NC$Total_Cases,NC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Northern Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")
```

Death rate = 0.0295

<h4>9. Western Cape</h4>
```{r}
WC = subset(mydata_sorted,Province == "Western Cape")
plot(WC$Total_Cases,WC$Total_Deaths,
     pch = 19,
     cex = 0.3,
     col = "#cc0000",
     main = "Western Cape:Total cases vs Total deaths",
     xlab = "Total cases",
     ylab = "Total deaths")

```

Death rate = 0.0336

<h1>4.Data Visualization using Tableau</h1>
<p>To view the dashboard on Tableau Public use the link below:</p>
<p>
https://public.tableau.com/app/profile/mufaro.chinake/viz/SouthAfricaCOVID19Analysis/Dashboard1
</p>
